---
title: "Chapter 14 Exercise 3"
author: "Carilli"
date: '`r format(lubridate::today(), "%B %d, %Y")`'
output: 
  html_document:
    toc: false
    toc_float: false
    df_print: paged
    theme: cerulean
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA,
                      options(digits = 3, scipen = 999))
library(tidyverse)
```

# Exercise {.tabset}


## 3 {.tabset}

```{r data-4, message=FALSE}
oly <- 
  here::here("data", "olympics_HW.csv") %>% 
  read_csv() %>% 
  janitor::clean_names()
```


### a

The model is $medals_i = \beta_0 + \beta_1 population_i + \epsilon_i$

```{r 3a}
oly %>% 
  lm(medals ~ population, .) %>% 
  jtools::summ()
```

### b

Consider the model $medals_i = \beta_0 + \beta_1 population_i + \beta_2 gdp_i + \epsilon_i$ 

```{r 3b}
oly %>%
  select(gdp, population) %>%
  cor(use = "pairwise.complete.obs")
```

We would expect gdp to have positive effect on medals.  Since GDP is negatively correlated with population and we expect GDP and the number of medals to positively related (*i.e.*, we expect $\beta_2>0$), we expect $\hat\beta_1$ to be understated.  

Correlation of X~1~ and X~2~|>0| Effect of omitted variable $\beta_2$ on Y~0~|<0
----|:---:|:---:|:----:|
\>0 |Overstate coefficient|No bias|Understate coefficient  
0| No bias| No bias|No bias  
<0|Understate coefficient|No Bias |Overstate coefficient

Table: Table 14.1 Effect of Omitting X~2~ on Coefficient Estimate of X~1~

### c

```{r 3c}
oly %>% 
  lm(medals ~ population + gdp, .) %>% 
  jtools::summ()
```

### d

```{r 3d}
oly %>%
  select(host, population) %>%
  cor(use = "pairwise.complete.obs")
```

### e

```{r 3e}
oly %>% 
  lm(medals ~ population + host, .) %>% 
  jtools::summ()
```

### f

```{r 3f}
oly %>% 
  lm(medals ~ population + gdp + host + elevation + temp, .) %>% 
  lm.beta::lm.beta()
```

### g

```{r 3g}
oly %>% 
  mutate(noisy_gdp = gdp + rnorm(1)) %>% 
  lm(medals ~ population + noisy_gdp + host + elevation + temp, .) %>% 
  lm.beta::lm.beta()
```

